Pre-/Post-Trikafta in CFRD

### 2022-10-04 Colleen Wood/Pre-post Trikafta/pre_post_trikafta_analysis.Rmd
Update pre_post_trikafta_analysis.Rmd

```{r setup, include=FALSE}
library(arsenal)
library(tidyverse)
library(lmerTest)
library(nlme)
library(broom.mixed)
library(segmented)
library(performance)
library(AGD)
library(rspiro)
library(lubridate)
library(plotly)
library(knitr)
knitr::opts_chunk$set(echo = FALSE,fig.height = 8,fig.width = 10)
if(Sys.info()["sysname"] == "Windows"){
  home_dir = "S:/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Linux"){
  home_dir = "~/UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Darwin"){
  home_dir = "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
}
source("~/GitHub/CF/get_genotype_class_severity.R")
knitr::opts_knit$set(root.dir = home_dir)
```

```{r include=FALSE}
# Import cleaned data from Casey
df = read.csv("./Colleen Wood/Prepost Triakfta/Data_Cleaned/analysis_data.csv",na.strings = "NA")
df$X = NULL
# Fill in MRN and CFF ID for everyone
df = df %>% 
  group_by(CFF.ID) %>% fill(MRN,.direction = "downup") %>% 
  group_by(MRN) %>% fill(CFF.ID,.direction = "downup")
# Assume that blanks for pancreatic and G tube indicate "no"
df$Hispanic.Latinx[is.na(df$Hispanic.Latinx)] = FALSE
df$Pancreatic.Status[is.na(df$Pancreatic.Status)] = "Sufficient"
df$G.tube.in.past.12.months[is.na(df$G.tube.in.past.12.months)] = "No"
# Convert dates
date_vars = c("DOB","Start","Date")
df[,date_vars] = lapply(df[,date_vars], function(d){
  parse_date_time(d,orders = c("mdy","ymd","dmy"),quiet = T)
})
# Fix incorrect century (WHY LUBRIDATE WHY?!)
df$DOB[which(df$DOB > lubridate::mdy("1/1/2022"))] = 
  df$DOB[which(df$DOB > lubridate::mdy("1/1/2022"))] - years(100)
# Time from start and filtering
df = df %>% group_by(CFF.ID) %>% fill(Start,.direction = "downup")
df = df %>% mutate(Time = as.numeric(difftime(Date,Start,units = "days"))) 
# Add UM (they did not provide dates, so need Time column)
um_a1cs = read.csv("./Colleen Wood/Prepost Triakfta/Data_Cleaned/um_a1cs.csv",na.strings = "")
um_bmi = read.csv("./Colleen Wood/Prepost Triakfta/Data_Cleaned/um_bmi.csv",na.strings = "")
um_demo = read.csv("./Colleen Wood/Prepost Triakfta/Data_Cleaned/um_demo.csv",na.strings = "")
um_ogtt = read.csv("./Colleen Wood/Prepost Triakfta/Data_Cleaned/um_ogtt.csv",na.strings = "")
um_pfts = read.csv("./Colleen Wood/Prepost Triakfta/Data_Cleaned/um_pfts.csv",na.strings = "")
um = full_join(um_a1cs,um_bmi)
um = full_join(um,um_ogtt)
um = full_join(um,um_pfts)
um = left_join(um,um_demo)
um = um %>% filter(!is.na(CFRD.Status))
um$Age = um$Age / 365.25
df = full_join(df,um)
# Filter and fill constants
df = df %>% group_by(CFF.ID) %>% fill(DOB,Start:CFRD.Diagnosis.Date,.direction = "downup") %>%
  ungroup() %>% filter(!is.na(Time),abs(Time) <= 730,!is.na(CFF.ID),
                       (!is.na(CFRD.Diagnosis.Date) | !is.na(CFRD.Status) | site == "UW")) %>%
  select(-MRN) %>% select(CFF.ID,site,DOB,Start,everything()) %>% rename(Site = "site")
df$Sex = as.factor(df$Sex)
levels(df$Sex) = c("F","M")
# BMI category
df$Age = pmax(as.numeric(difftime(df$Date,df$DOB,units = "days"))/365.25,df$Age,na.rm = T)
# Treat adults as age = 20
df$Age_for_BMI = ifelse(df$Age > 20,20,df$Age)
df$BMIz = y2z(df$BMI,x = df$Age_for_BMI,sex = df$Sex,ref=nl4.bmi)
df$BMIz[is.infinite(df$BMIz)] = NA
df$BMI.Percentile = pnorm(df$BMIz)*100
df$Weightz = y2z(df$Weight,x = df$Age_for_BMI,sex = df$Sex,ref=nl4.wgt)
df$BMI_group = apply(df,1,function(r){
  age = as.numeric(r["Age"])
  if(!is.na(age) & age > 19){
    bmi = as.numeric(r["BMI"])
    group = cut(bmi,c(-Inf,18.5,25,30,35,40,Inf),
        labels = c("Under Weight","Normal Weight","Overweight","Obese (Class I)",
                   "Obese (Class II)","Obese (Class III+)"),right = F)
  } else if (!is.na(age) & age <= 19){
    z = as.numeric(r["BMIz"])
    perc = pnorm(z)
    group = cut(perc,c(-Inf,5,85,95,Inf),
        labels = c("Under Weight","Normal Weight","Overweight","Obese (Class I)"),right = F)
  } else {group = NA}
  return(as.character(group))
})
df$BMI_group = as.factor(df$BMI_group)
df$BMI_group = ordered(df$BMI_group,c("Under Weight","Normal Weight","Overweight",
                                      "Obese (Class I)","Obese (Class II)","Obese (Class III+)"))
# Get genotype severity
df$Genotype1_sev = get_cf_genotype_class_severity(df$Genotypes1)
df$Genotype2_sev = get_cf_genotype_class_severity(df$Genotypes2)
# PFTs
old <- c("Caucasian","Black","Other")
new <- c(1,2,5)
df$ppFEV = pctpred_GLI(df$Age,df$Height/100,gender = as.numeric(df$Sex),
            ethnicity = new[match(df$Race, old)],
            FEV1 = df$FEV1)
df$ppFVC = pctpred_GLI(df$Age,df$Height/100,gender = as.numeric(df$Sex),
            ethnicity = new[match(df$Race, old)],
            FEV1 = df$FVC)
# T1D duration
df$CFRD.Diagnosis.Date = mdy(df$CFRD.Diagnosis.Date)
df$T1D.Duration = as.numeric(difftime(df$Date,df$CFRD.Diagnosis.Date,units = "days"))/365.25
# Before or after start
df$Before_After = cut(df$Time,c(-Inf,0,Inf),labels = c("Before","After"),right = T)
# Sort
df = df %>% arrange(Site,CFF.ID,Date) %>% distinct()
# Write sorted DF
write.csv(df, file = "./Colleen Wood/Prepost Triakfta/Data_Cleaned/analysis_data_with_UM.csv",
          row.names = F,na = "")

@@

```{r}
# Most recent before/after - limit to 2 years before start
# Want the value closest to start date before start, and most recent with respect to now from start
# (i.e. the longest time after start)
most_recent = df %>% group_by(CFF.ID,Site,Before_After) %>%
  filter(Time > -2*365) %>%
  mutate(time_new = max(Time) - Time) %>%
  arrange(time_new) %>% 
  summarise(Age = paste0(Date[which(Age == first(na.omit(Age)))],": ",round(first(na.omit(Age)),1)),
            BMI.Percentile = paste0(Date[which(BMI.Percentile == first(na.omit(BMI.Percentile)))],
                                    ": ",round(first(na.omit(BMI.Percentile)),1)),
            Weight = paste0(Date[which(Weight == round(first(na.omit(Weight)),1))],
                            ": ",first(na.omit(Weight))),
            T1D.Duration = paste0(Date[which(T1D.Duration == first(na.omit(T1D.Duration)))],
                                  ": ",round(first(na.omit(T1D.Duration)),1)),
            G.tube.in.past.12.months = first(na.omit(G.tube.in.past.12.months)),
            Pancreatic.Status = first(na.omit(Pancreatic.Status)),
            Genotype1_sev = first(na.omit(Genotype1_sev)),
            Genotype2_sev = first(na.omit(Genotype2_sev)),
            BMIz = paste0(Date[which(BMIz == first(na.omit(BMIz)))],
                          ": ",round(first(na.omit(BMIz)),1)),
            FEV1 = paste0(Date[which(FEV1 == first(na.omit(FEV1)))[1]],
                          ": ",round(first(na.omit(FEV1)),1)),
            FVC = paste0(Date[which(FVC == first(na.omit(FVC)))],
                         ": ",round(first(na.omit(FVC)),1)),
            HbA1c = paste0(Date[which(HbA1c == first(na.omit(HbA1c)))],
                           ": ",round(first(na.omit(HbA1c)),1)),
            .groups = "keep") %>%
  arrange(Site,CFF.ID,Before_After)
most_recent[most_recent == ": NA"] = NA
most_recent[most_recent == "NA: NA"] = NA

# # Check
# t = df %>% select(CFF.ID,Site,Date,Time,Before_After,BMIz,ppFEV,ppFVC) %>%
#   group_by(CFF.ID,Site,Before_After) %>%
#   arrange(Site,CFF.ID,Before_After)
# Filter those without CFRD, and those without pre AND post
keep = most_recent %>% filter(!is.na(HbA1c)) %>%
  group_by(CFF.ID) %>% summarise(n = n()) %>%
  filter(n == 2) %>% pull(CFF.ID)
cfrd = unique(df$CFF.ID[!is.na(df$CFRD.Status) | !is.na(df$CFRD.Diagnosis.Date)])
keep = unique(keep,cfrd)
most_recent = most_recent %>% filter(CFF.ID %in% keep)
```

@@

@@

# Piecewise Models

```{r}
# Piecewise model function
piecewise = function(outcome = "HbA1c",data = df,simple = F){
  # Add spline column
  data$Change = ifelse(data$Time < 0,0,1)
  data$Change = data$Change * data$Time
  # Model
  f = as.formula(paste0(outcome,"~Time + Change + (1|CFF.ID)"))
  mod = lmer(f,data = data)
  res = tidy(mod,"fixed")
  res$effect = NULL
  # Model check if necessary
  check = check_model(mod)
  # Coordinates for model plotting
  x1_start = -365
  y1_start = res$estimate[1] + res$estimate[2]*x1_start
  x1_end = 0
  y1_end = res$estimate[1] 
  x2_start = 0
  y2_start = res$estimate[1] 
  x2_end = 365
  y2_end = res$estimate[1] + (res$estimate[2]+res$estimate[3])*x2_end
  # Plot
  p = ggplot(data,aes_string(x = "Time",y = outcome,group = "CFF.ID",color = "Site")) +
    geom_point(alpha = 0.5) + geom_line(alpha = 0.2) +
    annotate("segment",x=x1_start,y=y1_start,xend=x1_end,yend=y1_end,color = "blue",size = 1) +
    annotate("segment",x=x2_start,y=y2_start,xend=x2_end,yend=y2_end,color = "blue",size = 1) +
    xlab("Days from Trikafta Start") + theme_bw()
  # Interactive plot and model results
  return(list(plot = ggplotly(p),table = res,check = check))
}
```

# Methods

For each outcome, we fit a piecewise linear mixed effects model with random intercept for each person and a "knot" at 0. This approach can be thought of as fitting two separate linear models, while ensuring that the two regression lines meet at Trikafta start. In the model results, "Slope Prior" indicates the slope prior to start of treatment (i.e. the average change in outcome for each "1 unit increase" in time). The "Change" term indicates the *change in slope* after medication start. To find the slope post-Trikafta, you add the "Slope Prior" and "Change" estimates.

The random intercept model was selected because it performed better (by AIC) than a model with random intercept and continuous AR1 correlation structure. Random slope models did not converge for some outcomes, so were not used for the sake of consistency between results.

# HbA1c

```{r results='asis'}
a1c = piecewise(outcome = "HbA1c",data = df)
ggplotly(a1c$plot)
kable(a1c$table,digits = 3)
```

`r if(class(a1c$seg)=="lme"){"Automated R package was not able to find a better breakpoint."}`

# BMI

Models did not converge with BMI converted to under weight, normal weight, etc. This is most likely due to lack of within-patient variability in terms of BMI group (i.e. people's BMIs aren't changing enough that they move from one group to another).

# Insulin

Most participants only have one observation of insulin dose, so these results would be the same as Maxie's previous reports.

# ppFEV

```{r results='asis'}
ppFEV = piecewise(outcome = "ppFEV",data = df)
ggplotly(ppFEV$plot)
kable(ppFEV$table,digits = 3)
```

# ppFVC

```{r results='asis'}
ppFVC = piecewise(outcome = "ppFVC",data = df)
ggplotly(ppFVC$plot)
kable(ppFVC$table,digits = 3)
```

# CGM Metrics

## Mean

```{r results='asis'}
meansg = piecewise(outcome = "average_sensor",data = df)
ggplotly(meansg$plot)
kable(meansg$table,digits = 3)
```

## Min

```{r results='asis'}
minsg = piecewise(outcome = "min_sensor",data = df)
ggplotly(minsg$plot)
kable(minsg$table,digits = 3)
```

## Max

```{r results='asis'}
maxsg = piecewise(outcome = "max_sensor",data = df)
ggplotly(maxsg$plot)
kable(maxsg$table,digits = 3)
```

## % Time > 120

```{r results='asis'}
over120 = piecewise(outcome = "percent_time_over_120",data = df)
ggplotly(over120$plot)
kable(over120$table,digits = 3)
```

## % Time > 140

```{r results='asis'}
over140 = piecewise(outcome = "percent_time_over_140",data = df)
ggplotly(over140$plot)
kable(over140$table,digits = 3)
```

## % Time > 200

```{r results='asis'}
over200 = piecewise(outcome = "percent_time_over_200",data = df)
ggplotly(over200$plot)
kable(over200$table,digits = 3)
```

## Mean excursions > 140

```{r results='asis'}
numover140 = piecewise(outcome = "avg_excur_over_140_per_day",data = df)
ggplotly(numover140$plot)
kable(numover140$table,digits = 3)
```

## Mean excursions > 200

```{r results='asis'}
numover200 = piecewise(outcome = "avg_excur_over_200_per_day",data = df)
ggplotly(numover200$plot)
kable(numover200$table,digits = 3)
```

## % Time < 70

```{r results='asis'}
under70 = piecewise(outcome = "percent_time_under_70",data = df)
ggplotly(under70$plot)
kable(under70$table,digits = 3)
```

## % Time < 60

```{r results='asis'}
under60 = piecewise(outcome = "percent_time_under_60",data = df)
ggplotly(under60$plot)
kable(under60$table,digits = 3)
```

## Average AUC per day

```{r results='asis'}
auc = piecewise(outcome = "average_auc_per_day",data = df)
ggplotly(auc$plot)
kable(auc$table,digits = 3)
```

## SD

```{r results='asis'}
sd = piecewise(outcome = "standard_deviation",data = df)
ggplotly(sd$plot)
kable(sd$table,digits = 3)
```

## MAGE

```{r results='asis'}
mage = piecewise(outcome = "r_mage",data = df)
ggplotly(mage$plot)
kable(mage$table,digits = 3)
```

>>>>

```{r setup, include=FALSE}
library(arsenal)
library(tidyverse)
library(lubridate)
source("~/GitHub/CF/get_genotype_class_severity.R")
variables = c("CFF ID","Site","Sex","Race","Genotypes1","Genotypes2","Pancreatic Status",
              "G-tube in past 12 months","Start","Date","Age")
```

```{r include=FALSE}
# CHCO
chco_a1c = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/chco_glycemic.csv",show_col_types = F)
chco_a1c$Date = parse_date_time(chco_a1c$Date,orders = "mdy")
chco_bmi = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/chco_bmi_pft.csv",show_col_types = F)
chco_bmi$Date = parse_date_time(chco_bmi$Date,orders = "mdy")
chco = full_join(chco_a1c,chco_bmi,by = c("MRN", "Date"))
# Fill CFF ID by MRN
chco= chco %>% group_by(MRN) %>% 
  fill(`CFF ID`,.direction = "downup") %>% ungroup() %>%
  filter(!is.na(`CFF ID`)) %>%
  select(-MRN)
# Add demographics
chco_reg = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/chco_registry.csv",show_col_types = F)
chco_reg = chco_reg %>% select(-MRN)
chco_reg$Start = parse_date_time(chco_reg$Start,orders = "mdy")
chco = left_join(chco,chco_reg,by = "CFF ID")
# Add Time
chco$Time = as.numeric(difftime(chco$Date,chco$Start,units = "days"))
```

```{r include=FALSE}
# Montana
mon_bmi = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/montana_bmi_pft.csv",show_col_types = F)
mon_bmi$Date = parse_date_time(mon_bmi$Date,orders = "mdy")
mon_a1c = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/montana_glycemic_registry.csv",show_col_types = F)
mon_a1c$Date = parse_date_time(mon_a1c$Date,orders = "mdy")
# Get start date and CFRD status
mon_starts = mon_a1c %>% select(`CFF ID`,`CFRD Status`,Start) %>% 
  distinct() %>% filter(`CFRD Status`=="CFRD with or without fasting hyperglycemia")
mon_starts$Start = parse_date_time(mon_starts$Start,orders = "mdy")
# Remove unnecessary columns
mon_a1c = mon_a1c %>% select(`CFF ID`,Date,HbA1c,`OGTT Fasting`,`OGTT Two Hour`)
# Merge and remove those without CFRD or start date
mon = full_join(mon_bmi,mon_a1c,by = c("CFF ID", "Date"))
mon= mon %>% filter(`CFF ID` %in% mon_starts$`CFF ID`)
mon = left_join(mon,mon_starts %>% select(`CFF ID`,Start), by = "CFF ID")
# Add Time
mon$Time = as.numeric(difftime(mon$Date,mon$Start,units = "days"))
```

```{r include=FALSE}
# UM
um_a1c = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/um_a1cs.csv",show_col_types = F)
um_a1c$Age = round(um_a1c$Age / 365.25,1)
um_bmi = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/um_bmi.csv",show_col_types = F)
um_bmi$Age = round(um_bmi$Age / 365.25,1)
um_ogtt = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/um_ogtt.csv",show_col_types = F)
um_ogtt$Age = round(um_ogtt$Age / 365.25,1)
um_pft = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/um_pfts.csv",show_col_types = F)
um_pft$Age = round(um_pft$Age / 365.25,1)
um_race = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/um_race.csv",show_col_types = F)
# Join
um = reduce(list(um_a1c,um_bmi,um_ogtt,um_pft),full_join)
um = left_join(um,um_race,by = "CFF ID")
```

```{r include=FALSE}
# UW
uw_bmi = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/uw_bmi.csv",show_col_types = F)
uw_bmi$Date = parse_date_time(uw_bmi$Date,orders = "mdy")
uw_glyc = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/uw_glycemic.csv",show_col_types = F)
uw_glyc$Date = parse_date_time(uw_glyc$Date,orders = "mdy")
uw_pft = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/uw_pft.csv",show_col_types = F)
uw_pft$Date = parse_date_time(uw_pft$Date,orders = "mdy")
uw_reg = read_csv("/Volumes/Documents/Work/Vigers/CF/Colleen Wood/Prepost Triakfta/Data_Cleaned/uw_registry.csv",show_col_types = F)
uw_reg$Start = parse_date_time(uw_reg$Start,orders = "mdy")
# Merge
uw = reduce(list(uw_bmi,uw_glyc,uw_pft,uw_reg),full_join)
```

```{r include=FALSE}
# Put everything together

@@

@@

```{r results='asis'}
t1 = tableby(~Age+BMI.Percentile+Weight+Sex+,pre_post[pre_post$Before_After == "Before",])
summary(t1)
```

@@
